### 2011-11-07 doc/pldi2012/highlevel.tex

The biggest challenge to implementing the high-level
runtime is to avoid the latencies associated with distributed
memory systems and deep memory hierarchies.  To aid
in solving this problem, the high-level runtime employs two useful 
scheduling techniques.  The first technique was proved
in Section \ref{sec:exec}: all scheduling decisions for a
task can be made locally in the context in which the
task was created.  The second technique is a {\em deferred execution model} 
that allows the runtime to hide the latencies associated
with moving data through deep memory hierarchies.  A
deffered execution model decouples the issuing of operations
from when they are actually performed.  This model is supported
by features in the low-level runtime.

@@

A deffered execution model however presents challenges to implementing
the semantics of the Legion programming model.  Deffered execution
does not necessarily imply that tasks will be executed in the order in
which they are scheduled, but only in the order in which dependences
are expressed.  To orchestrate the execution of Legion
programs in this environment the high-level runtime is architected
in a manner similar to a pipelined out-of-order hardware processor, but at
a coarser granularity with tasks corresponding to instructions and
regions corresponding to registers.  There is a pipeline of stages to 
executing a task which allows the runtime to extract
as much task-level parallelism as possible from an application whithin
the constraints of the Legion programming model.  To illustrate
the implementation of the high-level runtime we now describe the
details of how each stage of task execution is performed.

\subsection{Dependence Analysis} 
\label{subsec:depanalysis}
The first step in the execution of a task is dependence analysis.  When
a task is registered with the high-level runtime the task call includes
information about both the logical regions that the task will require
as well as the read-write privileges and coherence properties
for each of the regions.  The runtime uses this information to perform
dependence detection with each of the previous tasks that have been registered
in the same context.  Dependence detection only needs to be performed at
the scope of an enclosing context as was proved in Section \ref{sec:exec}.

We initially describe how dependences are detected from logical regions
in Section \ref{subsec:cohdep}.  We then describe the algorithm that
is used to perform dependence detection in Section \ref{subsec:depdetect}.
Section \ref{subsec:dataflow} describes how data is versioned correctly
during deffered execution.  For concreteness we describe how dependence
analysis is performed for the circuit simulation from Section \ref{sec:ex}.
%When performing dependence analysis the runtime relies on its knowledge about
%the relationships between logical regions and partitions.  We first describe
%the the components to dependence analysis in sections \ref{subsec:regiontree},
%\ref{subsec:cohdep}, and \ref{absinsts}.  We then describe the algorithm for performing 
%dependence analysis on this data structures in section \ref{subsec:depdetect}.
%\subsubsection{Region Trees}
%\label{subsec:regiontree}
%To detect dependences between the tasks, the runtime leverages its knowledge
%about logical regions and their partitions.  To create a new logical region
%or partition, the application must invoke the corresponding call in the
%runtime.  As these calls occur, the runtime constructs a data strcture
%called a {\em region tree}.  A region tree describes the relationship between
%logical regions and partitions.  A region tree contains two types of nodes:
%region nodes and partition nodes.  Every region tree is rooted with a region
%node and alternates between region and partition nodes at each corresponding level.
%A region node represents a specific logical region and tracks the set of
%partitions of that logical region.  Similarly a partition node represents 
%a specific partitioning of a logical region and keeps track of the logical regions 
%which are subregions of the partition.  
\subsubsection{Dependencies From Coherence and Privileges}
\label{subsec:cohdep}
To detect dependencies between any two tasks using a logical region, we leverage
the information contained in the privileges and cohrence properties provided by the
task call.  Dependency checks are performed pair-wise between regions
to determine whether one task depends on the other.  When checking for
dependencies we first examine the region privileges.  If both tasks have reads privileges, then 
there is never a dependence.  If at least one task has write privileges we use table
\ref{tab:depsdetect} to determine whether there is a dependence.
%In this table, {\em Dependence} indicates that there will always be a dependence between
%regions while {\em None} indicates that there will never be a dependence regardless
%of mapping decisions.  In the case of {\em Same Instance} there is a dependence contingent
%upon a mapping decision.  If both tasks map to the same physical instance (described in 
%more detail in section \ref{subsec:instmang}) then there is no dependence, which if
%they map to different instances there is a dependence as the system will have to generate
%a copy between the two tasks running.
For the case of {\tt Contingent} there are two special cases that we handle.
One special case is Write-After-Read (WAR) dependencies.  For the WAR case, if neither coherence property
is {\em Exclusive} then we can ensure that different physical instances
are used to avoid an anti-dependence and increase parallelism.  If either property is {\em Exclusive} then the semantics
of the programming model mandate that we detect a dependence and serialize the task.

Another special case occurs if there is an {\tt Atomic} writer and either {\tt Simultaneous}
or {\tt Relaxed} reader.  In this case it is acceptable for the tasks to run in
parallel as it is valid for the reader to see updates form the writer, and the
reader will not violate the atomic access semantics of the writer.
\begin{center}
\begin{table}
\small
\begin{tabular}{|c|c|c|c|c|} \hline
             & Exclusive & Atomic   & Simultaneous & Relaxed \\ \hline 
Exclusive    & Dep & Dep & Dep & Dep \\ \hline
Atomic       & Dep & Same & Cont & Cont \\ \hline
Simultaneous & Dep & Cont & Same & None \\ \hline
Relaxed      & Dep & Cont & None & None \\ \hline
\end{tabular}
\caption{Dependence detection cases when at least one access has a write privilege. {\tt Dep}
indicates a dependence is detected.  {\tt Same} indicates a dependence is detected
only if the task uses different instances.  {\tt Cont} means that the dependence
is contingent upon the permissions.  {\tt None} means there will never
be a conflict.\label{tab:depsdetect}}
\end{table}
\end{center}

@@

\subsubsection{Dependence Detection}
\label{subsec:depdetect}
When a task is registered, the runtime
will perform dependence detection for every region that the task requests with
prior tasks that have been registered in the same context.  Since
tasks are registered in program order we will always perform dependence detection
in program order which guarantees correctness even if tasks are executed out-of-order.  
Performing dependence detection in program order is annalogous to how an out of order
processor works.  Out-of-order processors load instructions in order and only after
analyzing their instructions and register dependencies do they execute instructions
out of order.  The regions that a task requires are annalogous
to the registers listed in an instruction and the goal of the high-level
runtime is to detect the same dependences and use them to enforce scheudling constraints
upon a task.  However, there is a catch to this analogy: in the Legion programming
model two different logical regions are not necessarily independent from each other.  In
order to perform dependence detection we rely on knowledge of the forest of regions to be
able to detect dependences between tasks which require different logical regions.

To perform dependence detection, we first find the {\em root region} for the 
requested region.  The root region is the ancestor region for which the parent task
has privliges.  Once we have found the root region for the requested region, we then proceed to walk
down the region tree from the root region to the node
for the requested region.  At each node, we determine if any potential dependences exist.
Determining whether dependencies exist involves checking information both about the current
node being visited as well as whether there are any potential dependences farther down
in the region tree.

When traversing a region node, we begin by checking if there are any tasks actively using
the node.  We determine this using the pair-wise dependence analysis described 
in Section \ref{subsec:cohdep} to detect dependences between the region we are looking for and the
region represented by the current node.  For each dependence that is found, we record that the previous
task must have completed prior to the start of the task we are registering.

After performing
dependence analysis we then check if there is an open partition which may contain additional
dependencies.  If the open partition is the partition we intend to traverse then we
simply continue the traversal.  However, if the partition we want is not open then we must
first {\em close} up any open partition.   The act of closing a partition 
involves traversing the region tree and recording
all active tasks as dependences.  The reason for closing a partition is that we have no disjointness
information about tasks that use different partitions and therefore to be safe must register
all task touching subregions as dependences.  After closing any open partitions, we
can then open the partition on our path and continue the traversal.

When we reach our target logical region we again perform the pair-wise dependence analysis.
We also close up any open partitions as they might contain tasks which could contain
a dependence.  If our task was dependent on all of the previous active tasks in a region
we say that our task was a {\em dominator} of all the previous active tasks.  We can
then clear the list of active tasks and replace it with the current task.  This is safe since
any later tasks that depend on the current task will have a transitive dependence 
through the dominator task.  If our task is not a dominator task, then we simply append
it to the list of active tasks making it possible for future task registrations to detect
it as a dependence.

When traversing partition nodes we perform a similar, but less expensive analysis.  There
are two cases to consider for partition nodes: disjoint partitions and aliased partitions.
In the case of disjoint partitions, we simply open the subregion that we intend to traverse
and then traverse it.  This is correct because we know by definition that any two 
subregions in a disjoint partition are disjoint and therefore a dependence can never exits
between them.

For aliased partitions we have to perform an additional dependence analysis.  At every
aliased partition node we maintain a list of tasks active in all of the subregions.  We
perform a pair-wise dependence analysis between every pair of tasks regardless of the subregion
they are accessing to see if a dependence could exist.  We do this regardless of the subregion
because we have no disjointness information in an aliased partition.  If there are any
potential dependences, we close up all the subregions and then open the subregion required.  If
no dependence was detected then we open the subregion we need, append the task to the list
of active tasks, and continue the traversal.  Allowing multiple open subregions in an aliased
partition is safe because we performed an analysis between all active tasks regardless of
whether the particular subregions were aliased or not.

>>>>

The biggest challenge in implementing the high-level
runtime is to avoid the latencies associated with distributed
memory systems and deep memory hierarchies.  To aid
in solving this problem, the high-level runtime employs two useful 
scheduling techniques.  The first technique, validated in 
Section~\ref{sec:exec}, is to make all scheduling decisions for a
task locally, i.e. in the context in which the
task was created.  The second technique is a {\em deferred execution model} 
that allows the runtime to hide the latencies associated
with moving data through deep memory hierarchies by decoupling the issuing
of operations
from when they are actually performed.  This model is supported
by features in the low-level runtime.

@@

A deferred execution model however presents challenges to implementing
the semantics of the Legion programming model.  Deferred operations
may not be executed in the order they were scheduled if no dependency
is expressed between them.  To orchestrate the execution of Legion
programs in this environment the high-level runtime uses an architecture
similar to that of a pipelined out-of-order hardware processor, but at
a coarser granularity.  Tasks can be though of as corresponding to instructions
and regions to registers.  The runtime uses a pipeline of task execution stages
to extract
as much task-level parallelism as possible from an application, while
preserving the guarantees of the Legion programming model.  To illustrate
the implementation of the high-level runtime we now describe the
details of how each stage of task execution is performed.

\subsection{Dependence Analysis} 
\label{subsec:depanalysis}
The first step in the execution of a task is dependence analysis.  Tasks are
registered with the high-level runtime in sequential program order.  The
registration lists the logical
regions to be accessed, specifying the needed privileges and the coherence.
The runtime uses this information to perform
dependence detection with each of the previous tasks that have been registered
in the same context (i.e. called by the same parent task).  These checks will
be sufficient to ensure correct program order, as was proved in Section~\ref{sec:exec}.

%We initially describe how dependences are detected from logical regions
%in Section \ref{subsec:cohdep}.  We then describe the algorithm that
%is used to perform dependence detection in Section \ref{subsec:depdetect}.
%Section \ref{subsec:dataflow} describes how data is versioned correctly
%during deferred execution.  For concreteness we describe how dependence
%analysis is performed for the circuit simulation from Section \ref{sec:ex}.

%When performing dependence analysis the runtime relies on its knowledge about
%the relationships between logical regions and partitions.  We first describe
%the the components to dependence analysis in sections \ref{subsec:regiontree},
%\ref{subsec:cohdep}, and \ref{absinsts}.  We then describe the algorithm for performing 
%dependence analysis on this data structures in section \ref{subsec:depdetect}.
%\subsubsection{Region Trees}
%\label{subsec:regiontree}
%To detect dependences between the tasks, the runtime leverages its knowledge
%about logical regions and their partitions.  To create a new logical region
%or partition, the application must invoke the corresponding call in the
%runtime.  As these calls occur, the runtime constructs a data strcture
%called a {\em region tree}.  A region tree describes the relationship between
%logical regions and partitions.  A region tree contains two types of nodes:
%region nodes and partition nodes.  Every region tree is rooted with a region
%node and alternates between region and partition nodes at each corresponding level.
%A region node represents a specific logical region and tracks the set of
%partitions of that logical region.  Similarly a partition node represents 
%a specific partitioning of a logical region and keeps track of the logical regions 
%which are subregions of the partition.  
\subsubsection{Dependencies From Coherence and Privileges}
\label{subsec:cohdep}
Conceptually, dependence detection with a previous task requires comparing the two sets of
logical regions being accessed.  For each logical reason from the previous task
that may overlap with a logical region needed by the new task, the privileges
and coherence modes are compared to determine whether a dependence exists.  If
only read privileges are in use, there is never a conflict, but if either or
both task needs write or reduction privileges, the coherence modes are compared
according to this table:

\vspace{2mm}
{\small
\begin{tabular}{c|cccc}
             & Exclusive & Atomic   & Simultaneous & Relaxed \\
\midrule
Exclusive    & Dep & Dep & Dep & Dep \\ 
Atomic       & Dep & Same & Cont & Cont \\
Simultaneous & Dep & Cont & Same & None \\
Relaxed      & Dep & Cont & None & None \\
\end{tabular}
}
\vspace{2mm}

An entry of {\tt Dep} means there is always a dependence, whereas {\tt None}
means that there is no dependency created by this region pair.  An entry of
{\tt Same} indicates a dependency that can be avoided if the two tasks are
able to use the same physical region.  Finally, an entry labeled {\tt Cont}
indicates a dependency that can be avoided if a single writer is using the
atomic coherence mode.  When a dependence is detected, the new task will
generally have to wait for the older task to complete before it can begin
execution.

%For each previously registered task, 
%To detect dependencies between any two tasks using a logical region, we leverage
%the information contained in the privileges and cohrence properties provided by the
%task call.  Dependency checks are performed pair-wise between regions
%to determine whether one task depends on the other.  When checking for
%dependencies we first examine the region privileges.  If both tasks have reads privileges, then 
%there is never a dependence.  If at least one task has write privileges we use table
%\ref{tab:depsdetect} to determine whether there is a dependence.

%In this table, {\em Dependence} indicates that there will always be a dependence between
%regions while {\em None} indicates that there will never be a dependence regardless
%of mapping decisions.  In the case of {\em Same Instance} there is a dependence contingent
%upon a mapping decision.  If both tasks map to the same physical instance (described in 
%more detail in section \ref{subsec:instmang}) then there is no dependence, which if
%they map to different instances there is a dependence as the system will have to generate
%a copy between the two tasks running.
%For the case of {\tt Contingent} there are two special cases that we handle.
%One special case is Write-After-Read (WAR) dependencies.  For the WAR case, if neither coherence property
%is {\em Exclusive} then we can ensure that different physical instances
%are used to avoid an anti-dependence and increase parallelism.  If either property is {\em Exclusive} then the semantics
%of the programming model mandate that we detect a dependence and serialize the task.

%Another special case occurs if there is an {\tt Atomic} writer and either {\tt Simultaneous}
%or {\tt Relaxed} reader.  In this case it is acceptable for the tasks to run in
%parallel as it is valid for the reader to see updates form the writer, and the
%reader will not violate the atomic access semantics of the writer.
%XS\begin{center}
%\begin{table}
%\small
%\begin{tabular}{|c|c|c|c|c|} \hline
%             & Exclusive & Atomic   & Simultaneous & Relaxed \\ \hline 
%Exclusive    & Dep & Dep & Dep & Dep \\ \hline
%Atomic       & Dep & Same & Cont & Cont \\ \hline
%Simultaneous & Dep & Cont & Same & None \\ \hline
%Relaxed      & Dep & Cont & None & None \\ \hline
%\end{tabular}
%\caption{Dependence detection cases when at least one access has a write privilege. {\tt Dep}
%indicates a dependence is detected.  {\tt Same} indicates a dependence is detected
%only if the task uses different instances.  {\tt Cont} means that the dependence
%is contingent upon the permissions.  {\tt None} means there will never
%be a conflict.\label{tab:depsdetect}}
%\end{table}
%\end{center}

@@

\subsubsection{Dependence Detection}
\label{subsec:depdetect}

Although the above algorithm for dependence detection is correct, the
high-level runtime is able to perform the detection much more efficiently with
the help of a custom data structure.

To perform dependence detection, the runtime will first find the {\em root region} for the 
requested region.  The root region is the ancestor region for which the parent task
has priviliges.  Once the root region for the requested region is found, 
the runtime can walk down the region tree from the root region to the node
for the requested region.  Each node contains a list of tasks that are
actively using the region corresponding to the node, potential dependencies are
discovered by comparing privileges and coherence modes with any tasks on that
list.

After check the list of tasks actively using the current node, the runtime
checks to see if one of the partitions of current node is {\em open}.  An
open partition indicates that there are some active tasks using subregions
of the current region.  No more than one partition of a node can be open
at any given time.  If the open partition is is along the path to the desired
region's node, traversal continues.  If the open partition lies along another 
path, or if the current node is the intended destination, the runtime must
{\em close} the currently active partition because it cannot guarantee
disjointness with the region the new task wishes to access.  To close a
partition in the region tree, the runtime must find all active tasks in that
subtree and consider them all to be dependencies for the task currently being
registered.  With the previously active partition closed, the runtime can 
then open up the desired partition and traverse further down the tree as
needed.

When the runtime traverses an open partition, its behavior depends on the kind
of partition being traversed.  If the partition is disjoint, no other subregion
can overlap with the target region, so the state of the other subregions can
be ignored in the traversal.  However, if the partition is aliases, any one
of those other subregions could overlap with the current partition, so
the runtime must close all of the open subregions (recording new dependencies
as it does so) before opening the subregion it intends to.

After the runtime reaches the target logical region, it performs dependence
checks against any active tasks listed for that region.  If a dependence is 
discovered, it is added to the current task's dependence list.  It can then
be removed from the region tree's active task list, as the current task (which
will be pushed onto the active task list) {\em dominates} the previous task.
In a case where the new task has dependencies with all existing active tasks
for the target node, that whole list is replaced by a single entry for the
new task.

### 2011-11-07 doc/pldi2012/lowlevel.tex

The purpose of the low-level runtime is to provide a machine-independent layer
for the high-level runtime.  This abstraction will provide portability to the Legion
programming model and the high-level runtime.  To achieve this goal the low-level
runtime needs to hide the ideosynchracies of the underlying machine.  At the same time, it needs to avoid 
straying too far from the native capabilities of the underlying hardware 
and software in order to obtain an efficient implementation.  The Legion
low-level runtime does this by abstracting the machine into a base set
of objects required to implement the Legion high-level.  These objects,
include processors, memories, regions, events, and deffered locks.

@@

\subsection{Low-Level Abstractions}
\label{subsec:lowobjects}
\subsubsection{Events}
\label{subsec:events}
The low-level runtime supports a deffered execution model to help hide
the latencies associated with communicating in a distributed memory
system.  The basic building block of the low-level runtime's deferred execution model
is the {\em event}.  An event is a placeholder for the completion of a deferred
operation.  When an asynchronous call is made to any of the low-level runtime objects, an event is 
immediately returned to the caller.  The caller may query whether the event has {\em triggered}
which happens when an even occurs.  The common use case for events is
to provide them as prerequisites to subsequent requests made of the runtime. 
These dependences cause the later requests to be automatically deferred until the previous
operations have completed.  By using events as dependences, the caller can
set up complicated, but coordinated chains of operations to be executed without
having to wait for the low-level runtime to execute them.
%These requests will
%also return events, and the caller can set up a complicated, but coordinated,
%set of operations to be performed by the low-level runtime with no babysitting
%required of the original caller.

@@

\subsubsection{Deferred Locks}
\label{subsec:defferedlocks}
Events are used to provide a deterministic ordering between operations, but in
some cases is overly restrictive.  To implement the more relaxed coherence
properties supported in Legion we need a less restrictive ordering semantics that
can result in better performance.  The traditional tool for this job is
the mutex (a.k.a. lock), and it is from the common reader/writer lock that
the low-level runtime's {\em deferred lock} is derived.  The key visible 
difference between a mutex and a deferred lock is that the request does not
block the caller's execution.  Many libraries offer nonblocking locks, but
they achieve this by returning without the lock if it is unavailable.  If
a deferred lock is currently unavailable, the request is guaranteed to be
granted at some time in the future (with no further action required of the 
caller).  That time in the future is described with an event.  Like all other
events, it can be given to other threads, or even stored
in a data structure in memory to be extracted by an arbitrary thread at a 
later time.  This dissociation of a held lock from the thread that
requested it is another important difference between the deferred lock and
existing mutexes.

@@

\subsubsection{Region Meta-Data Objects}
\label{subsec:regionmeta}
Application-level regions are broken into four inter-related pieces in the
low-level runtime:
\begin{description}

@@

\subsubsection{SMP Implementation}
\label{subsec:smpimpl}
The SMP implementation of the low-level runs on any SMP computer that supports 
POSIX threads.  A single memory object is created for the shared memory, and 
a processor object is created for
each CPU core in the system.  Events are 
mutex-protected data structures that maintain a list of callbacks to call when
the corresponding event is triggered.  Locks are similar, except that the list
is of pending lock requests to attempt when the previous lock is released.

>>>>

The low-level runtime is a machine-independent layer that provides portability
to the high-level runtime and consequently the Legion programming model.
The low-level runtime hides the ideosynchracies of the underlying machine while
at the same time not straying 
too far from the native capabilities of the underlying hardware.  
The machine abstraction that the Legion low-level runtime presents is a base set
of objects.  These objects,
include processors, memories, regions, events, and deffered locks.

@@

\subsection{Low-Level Abstractions}
\label{subsec:lowobjects}
\subsubsection{Events}
\label{subsec:events}
The low-level runtime supports a deffered execution model to hide
the latencies associated with communicating in a distributed memory
system.  The basic building block of the low-level runtime's deferred execution model
is the {\em event}.  An event is a placeholder for the completion of a deferred
operation.  When an asynchronous call is made to any of the low-level runtime objects, an event is 
immediately returned to the caller.  The caller may query whether the event has {\em triggered}
which happens when the operation corresponding to the event completes.  The common use case for events is
to provide them as prerequisites to subsequent requests made of the runtime. 
These dependences cause the later requests to be automatically deferred until the previous
events have triggered.  By using events as dependences, the caller can
set up complicated, but coordinated chains of operations to be executed without
having to wait for the low-level runtime to execute them.
%These requests will
%also return events, and the caller can set up a complicated, but coordinated,
%set of operations to be performed by the low-level runtime with no babysitting
%required of the original caller.

@@

\subsubsection{Deferred Locks}
\label{subsec:defferedlocks}
Events are used to provide a deterministic ordering between operations, but in
some cases are overly restrictive.  To implement the more relaxed coherence
properties supported in Legion we need a less restrictive ordering semantics that
can result in better performance.  The traditional tool for this job is
the mutex (a.k.a. lock), and it is from the common reader/writer lock that
the low-level runtime's {\em deferred lock} is derived.  The key visible 
difference between a mutex and a deferred lock is that the request does not
block the caller's execution.  Many libraries offer nonblocking locks, but
they achieve this by returning without the lock if it is unavailable.  If
a deferred lock is currently unavailable, the request is guaranteed to be
granted at some time in the future (with no further action required of the 
caller).  That time in the future is described with an event.  Like all other
events, it can be given to other threads, or even stored
in a data structure in memory to be extracted by an arbitrary thread at a 
later time.  This dissociation of a held lock from the thread that
requested it is another important difference between the deferred lock and
existing mutexes.

@@

\subsubsection{Regions}
\label{subsec:regionmeta}
Application-level regions are broken into four inter-related pieces in the
low-level runtime:
\begin{description}

@@

\subsubsection{SMP Implementation}
\label{subsec:smpimpl}
The SMP implementation of the low-level runs on any SMP machine that supports 
POSIX threads.  A single memory object is created for the shared memory, and 
a processor object is created for
each CPU core in the system.  Events are 
mutex-protected data structures that maintain a list of callbacks to call when
the corresponding event is triggered.  Locks are similar, except that the list
is of pending lock requests to attempt when the previous lock is released.
